A Large-scale Study of Polymorphism in R

### 2019-04-02 paper/main.tex
changes all over

\newcommand{\attr}[2]{\ensuremath{#1\,({\mathit #2})}\xspace}
\renewcommand{\to}{\ensuremath{\rightarrow}\xspace}
\newcommand{\D}{\ensuremath{\small\vec{\mathtt D}}\xspace} % Double
\newcommand{\I}{\ensuremath{\small\vec{\mathtt I}}\xspace} % Integer
\renewcommand{\C}{\ensuremath{\small\vec{\mathtt C}}\xspace} % Character
\renewcommand{\L}{\ensuremath{\small\vec{\mathtt L}}\xspace} % Logical
\newcommand{\R}{\ensuremath{\small\vec{\mathtt R}}\xspace} % Raw
\newcommand{\X}{\ensuremath{\small\vec{\mathtt X}}\xspace} % Complex
\newcommand{\Y}{\ensuremath{\small\vec{\mathtt Y}}\xspace} % Symbol
\newcommand{\sY}{\ensuremath{\small{\mathtt Y}}\xspace} % Symbol
\newcommand{\sS}{\ensuremath{\small{\mathtt S}}\xspace} % S4
\newcommand{\sF}{\ensuremath{\small{\mathtt F}}\xspace} % Closure
\newcommand{\sE}{\ensuremath{\small{\mathtt E}}\xspace} % Env
\renewcommand{\R}{\ensuremath{\small\vec{\mathtt R}}\xspace} % Raw
\newcommand{\sN}{\ensuremath{\small{\mathtt N}}\xspace}     % Null
\renewcommand{\l}{\ensuremath{\small\underline{\mathtt ?}}\xspace}     % List
\newcommand{\sD}{\ensuremath{\small{\mathtt D}}\xspace} % Double
\newcommand{\sI}{\ensuremath{\small{\mathtt I}}\xspace} % Integer
\newcommand{\sC}{\ensuremath{\small{\mathtt C}}\xspace} % Character
\newcommand{\sL}{\ensuremath{\small{\mathtt L}}\xspace} % Logical
\newcommand{\sX}{\ensuremath{\small{\mathtt X}}\xspace} % Complex
\newcommand{\sR}{\ensuremath{\small{\mathtt R}}\xspace} % Raw
\newcommand{\ANY}{\ensuremath{\small{\mathtt ?}}\xspace}     % Any
\newcommand{\lT}[1]{\ensuremath{\small\underline{\mathtt #1}}\xspace}     % Any

@@

In R all arguments are passed as promises, and unused arguments will be
unevaluated.  Our tracer does still collect information on these arguments,
as whether or not they were used doesn't change the fact that they were in
fact passed.  In some cases, arguments aren't even intended to be evaluated.
For example, consider the {\tt magrittr} package, which implements the
Tidyverse's pipe functions.  Refer to Figure~\ref{fig:magpipeex} for a
translation of a call to {\tt magrittr}'s pipe ({\tt \%>\%}) function.

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
mul_by <- function(x, y) { x * y }

4 %>% mul_by(3) === `%>%`(lhs = 4, rhs = mul_by(3)) === mul_by(4, 3)
\end{lstlisting}}\caption{Example and rough translation of \%$>$\% call.}\label{fig:magpipeex}\end{figure}

Our tracer evaluates function arguments, and here \code{rhs = mul_by(3)},
but that's erroneous code (you can't call {\tt mul\_by} without supplying
two arguments).  In this situation, our tracer will still register the
entire call, but flag the {\tt rhs} argument to indicate to later stages of
the analysis that there was an issue evaluating that argument.

For each function of each package, we consolidate all recorded observations
into a function signature. The details of how consolidation is parameterized
by the definition of a potential type system for R and will be discussed in
the next section.

@@

We will need a notion of polymorphism, and thus a notion of type, when
undertaking this analysis.  R is unityped, and the only notion of
polymorphism in the language is its use of dynamic dispatch on the class of
an argument.  For our purposes, we will define a notion in type roughly in
line with R's \code{typeof} reflection function, in some cases augmented by
information obtained by \code{class} and \code{attributes} reflection.  We
start by defining what it means for an {\it argument} to be polymorphic.

\begin{itemize}
\item an argument is {\it polymorphic in type} if it has been called with at
  least two distinct types according to our definition of type.
\item an argument is {\it polymorphic in class} if it has been called with
  at least two distinct classes according to the \code{class} function.
\item attributes are a bit more tricky.  We can't simply count the number of
  attributes that have inhabited an argument, since values can contain
  arbitrarily many attributes (and having several attributes isn't
  polymorphism---think records).  Instead, we define the {\it attribute
    pattern} of a value to be the set of all attribute name and type pairs
  in the attribute list of a value (obtained via \code{attributes}).  And so
  an argument is {\it polymorphic in attribute} if it has been called with
  at least two distinct attribute patterns.
\end{itemize}

The main ingredient of our types is the {\tt typeof} a value, and there is
an important distinction to be made between the result of \code{typeof} and
that of \code{class} and \code{attributes}.  In R, the \code{typeof} a value
cannot change, and this is not the case with class and attributes.  As
mentioned in \AT{an earlier section}, the class of values can be easily
redefined by assigning to the value's class attribute, and similarly
attributes can be dynamically added, removed, or changed.

@@

\subsection{Granularity of Type Information}

When constructing a type for a value, we begin with the {\tt typeof} the
value.  This alone paints an incomplete picture of the value, though,
especially when it comes to vectors (recall that \code{typeof(5) ==
  typeof(c(5, 5)) == "double"}).  In some cases, the class or attributes of
a value might enlighten us, but in many cases we can be a little more
precise.

To be as specific as possible, our analysis collects additional information
to create a more specific type than the result of any of R's reflection
functions.  The details of the information we collect follows.

\begin{itemize}
\item if the analysis encounters \code{ typeof(x) == "double"}, it will look
  at the length of {\tt x} to determine if {\tt x} is a scalar or a vector,
  generating the annotation {\tt \dbls} or {\tt \dblv} accordingly.  This is
  also true for integer, complex, raw, logical, and character {\tt typeof}s
  (i.e., the primitive vectorized data types).  A notable imperfection here
  is is that unit length vectors will appear as scalars, so later in our
  analysis we ensure that argument signatures containing both scalar/X and
  vector/X are collapsed down to contain only vector/X.

\item if the analysis encounters \code{typeof(x) == "list"}, we collect type
  information on all list elements in order to ascribe a general type to the
  contents of the list.  To avoid undue slowdowns, we only collect the
  \code{typeof} the contents, and ascribe type {\tt any} if two or more {\tt
    typeof}s are observed.

\item if the analysis encounters \code{NA}, we ascribe a unique {\tt
  raw\_NA} type.  In R, \code{NA} inhabits all types, but \code{typeof(NA)
  == "logical"}.  The analysis deals with these {\tt raw\_NA}s at a later
  stage.

\item if the analysis encounters a matrix (i.e. the value has matrix class
  and has a dims and optionally dimnames attribute), we ascribe the type
  {\tt matrix}.

\item similarly, if it encounters a data frame (i.e., class {\tt data.frame}
  with appropriate attributes), we ascribe the type {\tt data.frame}.
\end{itemize}

%
%
%
\section{Examples of Polymorphism}
\label{sec:polyex}

>>>>

\newcommand{\attr}[2]{\ensuremath{#1_{\mathtt{#2}}}\xspace}
\renewcommand{\to}{\ensuremath{\rightarrow}\xspace}
\newcommand{\D}{\ensuremath{\small\vec{\mathtt D}}\xspace} % Double
\newcommand{\I}{\ensuremath{\small\vec{\mathtt I}}\xspace} % Integer
\renewcommand{\C}{\ensuremath{\small\vec{\mathtt C}}\xspace} % Character
\renewcommand{\L}{\ensuremath{\small\vec{\mathtt L}}\xspace} % Logical
\newcommand{\R}{\ensuremath{\small\vec{\mathtt R}}\xspace} % Raw
\newcommand{\X}{\ensuremath{\small\vec{\mathtt X}}\xspace} % Complex
\newcommand{\Y}{\ensuremath{\small\vec{\mathtt Y}}\xspace} % Symbol
\newcommand{\sY}{\ensuremath{\small{\mathtt Y}}\xspace} % Symbol
\newcommand{\sS}{\ensuremath{\small{\mathtt S}}\xspace} % S4
\newcommand{\sF}{\ensuremath{\small{\mathtt F}}\xspace} % Closure
\newcommand{\sE}{\ensuremath{\small{\mathtt E}}\xspace} % Env
\renewcommand{\R}{\ensuremath{\small\vec{\mathtt R}}\xspace} % Raw
\newcommand{\sN}{\ensuremath{\small{\mathtt N}}\xspace}     % Null
\renewcommand{\l}{\ensuremath{\small\underline{\mathtt ?}}\xspace}     % List
\newcommand{\sD}{\ensuremath{\small{\mathtt D}}\xspace} % Double
\newcommand{\sI}{\ensuremath{\small{\mathtt I}}\xspace} % Integer
\newcommand{\sC}{\ensuremath{\small{\mathtt C}}\xspace} % Character
\newcommand{\sL}{\ensuremath{\small{\mathtt L}}\xspace} % Logical
\newcommand{\sX}{\ensuremath{\small{\mathtt X}}\xspace} % Complex
\newcommand{\sR}{\ensuremath{\small{\mathtt R}}\xspace} % Raw
\newcommand{\ANY}{\ensuremath{\small{\mathtt ?}}\xspace}     % Any
\newcommand{\lT}[1]{\ensuremath{\small\underline{\mathtt #1}}\xspace}     % Any

@@

In R all arguments are passed as promises, and unused arguments will be
unevaluated.  Our tracer does not collect information on arguments that were
not evaluated as they may not be valid expressions.  Consider the {\tt
  magrittr} package, which implements the pipe function that performs a kind
of currying.

{\small\begin{lstlisting}[style=R]
mul <- function(x,y)  x*y
4 %>% mul(3) 
\end{lstlisting}}

The call to the pipe desugars to \code{`\%>\%`(lhs=4, rhs=mul(3))} where the
\code{rhs} is not a valid expression. It is treated as code that will
manipulated to add one argument (the \code{lhs}).

To be as specific as possible, our analysis collects shape information about
arguments and return values. The details of the information we collect follows.

\begin{itemize}
\item If the analysis encounters a primitive vector, say
  \code{typeof(x)=="double"}, it records the length of {\tt x} to determine
  if it can be treated as a scalar.
\item For a list, \code{typeof(x)=="list"}, shape information on all
  elements is used to describe the list's shape.  To avoid undue slowdowns,
  we only collect the content's \code{typeof}, and ascribe \ANY (any) if the
  elements are not of the same kind.k
\item If the analysis encounters a scalar \code{NA}, we ascribe a unique
  NULL type.  In R, \code{NA} inhabits all types, but for a scalar NA,
  \code{typeof(NA)=="logical"}. Scalar NAs are used as uninitialized values.
\item For a matrix (i.e. the value has matrix class and has a dims and
  optionally dimnames attribute), we ascribe the shape \attr{\tt T}{mat}
  where {\tt T} is the primitive element type.
\item For a data frame (i.e., class {\tt data.frame} with appropriate
  attributes), we ascribe the shape \attr{\l}{df}.
\end{itemize}

The infrastructure memoizes emitted signatures to avoid writing multiple
identical signatures for the same function.

@@

For a meaningful notion of polymorphism, it is necessary to settle on a set
of basic types for R. There are several possible choices which lead to
increasingly more expressive type systems. The simplest notion of types
aligns with R basic types. Richers notions incorporate attributes and the
various object systems built on top of those. For concreteness we will pick
out three different sets of types.

\begin{itemize}
\item {\bf L0}: This level includes vectors of primitive types (integers \I,
  double \D, character \C, logical \L, raw \R, complex \X), lists of any
  type (\l), and scalar values (environment \sE, closures \sF, symbols \sS
  and null \sN). We add the type \ANY for any type.
\item {\bf L1}: The second level enhances {\bf L0} with scalar values for
  primitive types (\sI, \sD, \sC, \sL, \sR, and \sX), type parametric lists
  (written \lT{T} for any type {\tt T}), data frames \attr{\l}{df}, and
  matrices \attr{\vec{\tt T}}{mat}.
\item {\bf L2}: The third level adds attributes lists...
\end{itemize}

In the following we say that an argument is {\it polymorphic in type at
  level L} if it has been called with at least two values that belong to
distinct types at level \emph{L}. This definition extends naturally to
classes, data frames, and matrices (as these are simple attributes).
Generalized attribute polymorphism is a bit trickier as values can have a
large and changing set of attributes.

@@

%
\section{Examples of Polymorphism}
\label{sec:polyex}
